---
title: "Data Analysis Report 1"
author: "Will Kramlinger"
date: "January 13, 2019"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Exploratory data analysis (EDA) was conducted on a data set comprising median
housing values of various Boston suburbs and potential predictors.
Correlations among predictors were examined, with particular attention paid to 
pairwise correlations with per capita crime rates, of which many were 
significant.  Commonalities amongst predictors were examined for suburbs with 
low median household values, as well as for suburbs which average a large number 
of rooms per dwelling.


## Data
The data set analyzed was the **Boston** data set included in the R package 
**MASS**.  The set includes observations for 506 suburbs across 14 variables. 
For variable definitions, please refer to <https://www.rdocumentation.org/packages/MASS/versions/7.3-47/topics/Boston>. 
The following shall presume the reader's familarity with the variable names and
their meanings, as they are used interchangeably.  A check for missing data 
within the set revealed no problems and no cleaning was deemed necessary.  A 
simple 5-number summary for each variable, along with each mean, was calculated 
as Table 1.

The project was an initial exploration of the data set intended as an exercise
in data visualization and basic descriptive statistics.  As the scope of 
this report is limited to EDA, an investigation of possible influential points 
was not undertaken.


## Analyses

*Correlation among predictors*

Correlation amongst the variables was investigated via scatterplot and
correlation matrices. A few of the most highly correlated linear relationships 
between predictor pairs are shown as Figs. 1a (**nox** vs. **age**, *r* = 0.731)
and 1b (**medv** vs **lstat**, *r* = -0.738).  While visual inspection of these
plots reveals clear trends, which are supported quantitively via the correlation
values, the trends do not appear to be strictly linear, but rather curvilinear;
this observation holds true for many of the variable pairs. Thus, for future 
regression analysis, transformation of variables will likely be necessary, with
transformations informed via residual analyses. Additionally, if a combination 
of these variables are used as predictors, a careful approach is advised as to 
avoid multicollinearity.

*Per capita crime rates*

Of particular interest in this analysis was the per capita crime rate (**crim**)
and its association with the other predictors in the data set.  While not 
reflected in the correlation matrix values, the variable demonstrated 
visible trends for nearly every instance that it was plotted pairwise 
against the other variables.   

Generally, 4 classes of trends were observed:  
1) The full range of crime rate values were found to be mostly included at one
particular level of the predictor; at all other levels of the predictor, the
crime rates were much more narrowly distributed and nearly equal to the minimum 
values of all crime rate values in the data set.  The subset of predictors for
which this trend was observed is discussed further below.  
2) Both the mean response and overall variation in the crime rate values 
increased as the predictor level either increased (positive correlation; **age**
, **lstat**) or decreased (negative correlation; **dist**, **medv**).  All of
these trends between variable pairs appeared more curvilinear than linear, 
akin to the previous discussion.  
3) The overall variation in the crime rate values was largest at 
intermediate values of the predictor, and grew smaller as
the predictor levels were moved further away from these intermediate values. In
other words, the curves resembled a unimodal distribution.  This trend was 
observed for predictors **nox** and **rm**.  
4) The overall variation in the crime rate values was smallest at 
intermediate values of the predictor, and grew larger as
the predictor levels were moved further away from these intermediate values. In
other words, the data resembed a bimodal distribution.  This trend was observed 
for predictor **black**.

The trend of particular note is the first class, which was observed when crime 
rate (**crim**) was plotted against each 
of **zn**, **indus**, **chas**, **rad**, **tax**, and **ptratio**.  A 
representative plot of this trend is shown as Fig. 2, where the majority 
of the variation in crime rate was observed at the specific predictor level 
**ptratio** = 20.2.  Other predictor levels where this pattern was observed 
included **zn** = 0, **indus** = 18.10, **chas** = 0, **rad** = 24, and **tax** 
= 666. Furthermore, it was found that of the suburbs in the top 25% (*n* = 127, 
**crim** > 3.667) in terms of crime rate, 120/127 = 94% of them shared *all* of 
these specific predictor values.  Further analysis revealed that these specific 
predictor levels, with the exception of **chas** = 0 (see below), are severely 
over-represented within these high-crime suburbs.  Comparison of these predictor 
levels with the summary in Table 1 reveals that these values are equal to either 
the 25th (**zn**, **chas**) or 75th (**indus**, **rad**, **tax**, **ptratio**)
percentiles for each particular variable in the data set, which suggests a 
large degree of commonality between these suburbs.

Of practical significance is the observation that this majority of suburbs with 
high crime rates had no residential lots over 25,000 sq ft (**zn** = 0), 
full-value property tax rate values which were near the highest observed in the 
overall data, a relatively high pupil:teacher ratio, and a high index of radial 
highway accessibility (research into whether a high index indicates high or low 
accessibility yielded no results).  Further investigation of these suburbs is 
warranted; a spatial analysis may be particularly useful.  Additionally, the 
plausibility of grouping these suburbs into one class should be considered for 
the sake of variable reduction in future analyses.  Please note that **chas** = 
0 (i.e., not on the Charles River) is specifically not mentioned for this 
particular analyses, as it is not considered practically 
meaningful. For **chas** = 0, the proportion of suburbs in the overall data set 
(471/506 = 0.931) was found to be not statistically different from the 
proportion found within the top 25% suburbs in terms of crime rate (120/127 =
0.945) at 99% confidence.

*Median values of owner-occupied homes*

As an initial step into understanding various predictors' relationships with the 
median household values (**medv**) of all Boston suburbs, the suburb with the 
lowest **medv** was identified.  Two suburbs were found to have this minimum 
value of **medv** = 5, namely, suburbs 399 and 406.  Each suburb's predictor 
levels and corresponding percentiles across the 13 other predictors are 
tabulated in Table 2.

Inspection of Table 2 reveals that the suburbs under analysis share common 
predictor levels across the majority of variables, with the exception of the
predictor **black**.  In terms of percentiles, the two suburbs differ from each
other by, at most, 8, with the exception of predictor **black** which differs 
by 66.  Recall the previous discussion about the first class of trends seen
for the variable **crim** and the associated predictor levels of **zn**,
**indus**, **chas**, **rad**, **tax**, and **ptratio**; these specific levels 
are also seen for suburbs 399 and 406, suggesting that suburbs with low median 
household values may be correlated with higher per capita crime rates.  

Additionally, the observed values for **nox**, **age**, **rm**, **dis**, and 
**lstat** are similar for the two suburbs. Specifically, the values of **nox**, 
**age**, and **lstat** for each suburb reside near the higher end of each 
variable's range; the values of **rm** and **dis** reside at the lower ends.  
Practically, this preliminary analysis suggests that suburbs with lower median 
household values may also be characterized via higher NO concentrations, older 
homes with relatively few rooms, a higher percentage of lower status population, 
and close proximities to Boston employment centers.

*Average number of rooms per dwelling*

An initial investigation into suburbs with higher levels of average number of
rooms per dwelling (**rm**) was conducted.  The proportion of suburbs within
the data set with values of **rm** > 7, 8 were found to be 12.6% and 2.6%,
respectively.

Comparisons of the mean response of variables for the groups
**rm** > 8 vs. **rm** < 8 were conducted via t-test at 95% confidence with a
null hypothesis of no difference; results are shown as Table 3.  Variables where 
statistically significant differences between group means were observed included 
**crim**, **indus**, **tax**, **ptratio**, **black**, **lstat**, and **medv**.  
Of these variables, the mean responses for predictors **black** and **medv** 
were found to be greater for the **rm** > 8 group; all other mean responses 
were found to be smaller for the **rm** > 8 group.  

## Plots and Tables

```{r Load, echo = FALSE}
# Load data set, check missing values, obtain 5-number summary
rm(list = ls())
library(MASS)
df <- Boston # Copy data for less typing and to maintain original set
```

Table 1: 5-number summary across all variables in the Boston data set
```{r Table1, echo = FALSE}
table1 <- summary(df)
names(dimnames(table1)) <- 
  list("", 
       "Table 1: 5 number summary and mean for each variable in Boston dataset")
table1
```


```{r fig1, echo = FALSE}
plot(df$nox ~ df$age, xlab = "Proportion", ylab = "Concentration (ppm)")
title(sub = "Fig. 1a")
plot(df$medv ~ df$lstat, xlab = '"Lower" Status (%)', ylab = "Value ($1000)")
title(sub = "Fig. 1b")
```

Fig. 1: (a) NO Concentration vs. Proportion of Units 
Built Before 1940 shows positive correlation. (b) 
Median Owner-Occupied Household Value vs. Proportion of Lower Status Population 
shows negative correlation.  Examples of typical trends seen between variable 
pairs in the data set with highest linear correlations; linear correlations 
likely to improve with variable transformations.


```{r fig2, echo = FALSE}
plot(df$crim ~ df$ptratio, xlab = "Ratio", ylab = "Crime Rate")
title(sub = "Fig. 2")
```

Fig. 2: Per capita crime rate vs. average pupil-teacher ratio. Example of trend 
seen when per capita crime rate is plotted pairwise against multiple other 
predictors, where most of the response variation is observed at one 
specific predictor level.  The variation is seen at **ptratio** = 20.2 in this
instance.


Table 2: Predictor values and associated percentiles for suburbs with the 
minimum median household value observed in dataset (**medv** = 5).  All 
predictor levels generally fall in the top or bottom 15% of the range of levels 
observedin full data set.  
```{r Pctiles, Table2, echo = FALSE}
pctiles_legend <- as.data.frame(matrix(ncol = 14, nrow = 100))
pctiles_min_medv <- as.data.frame(matrix(ncol = 14, nrow = 2))
min_medv_index <- which(df$medv == min(df$medv))
colnames(pctiles_legend) <- colnames(df)
colnames(pctiles_min_medv) <- colnames(df)
for (i in 1:14) {
  for(k in 1:2) {
    pctiles_legend[ , i] <- quantile(df[ , i], seq_len(100) * 0.01)
    pctiles_min_medv[k, i] <- findInterval(df[min_medv_index[k], i], 
                                            pctiles_legend[ , i])
  }
}
values_min_medv <- data.frame(rbind(df[399, ], df[406, ]))
table2 <- data.frame(rbind(round(values_min_medv[1, ], 1), 
                           round(pctiles_min_medv[1, ], 0),
                           round(values_min_medv[2, ], 1),
                           round(pctiles_min_medv[2, ], 0)))
row.names(table2) <- c("Suburb 399 - Values", "Suburb 399 - Percentiles",
                       "Suburb 406 - Values", "Suburb 406 - Percentiles")
table2[, -14]
```


Table 3: Statistics, p-values, and 95% CIs generated from 2-sample, 2-sided 
t-tests comparing groups of suburbs with more or less than 8 rooms per dwelling, 
with null hypothesis of no difference.
```{r Table3, echo = FALSE}
# Calculate percentiles for suburbs averaging more than 8 rooms/dwelling
pctiles_morethan8 <- as.data.frame(matrix(ncol = 14, nrow = 13))
morethan8_index <- which(df$rm > 8)
colnames(pctiles_morethan8) <- colnames(df)
for (i in 1:14) {
  for(k in 1:13) {
    pctiles_morethan8[k, i] <- findInterval(df[morethan8_index[k], i], 
                                            pctiles_legend[ , i])
  }
}
values_morethan8 <- df[which(df$rm > 8), ]
values_lessthan8 <- df[-which(df$rm > 8), ]
t_test_values <- as.data.frame(matrix(nrow = 4, ncol = 14))
rownames(t_test_values) <- c("t_statistic", "p-value", "95% CI, Low", 
                             "95% CI, High")
colnames(t_test_values) <- colnames(df)
for(i in 1:14) {
  p <- t.test(values_morethan8[ , i], values_lessthan8[ , i], conf.level = 0.95)
  t_test_values[1, i] <- round(p$statistic, 2)
  t_test_values[2, i] <- round(p$p.value, 3)
  t_test_values[3, i] <- round(p[[4]][1], 2)
  t_test_values[4, i] <- round(p[[4]][2], 2)
}
t_test_values[ , c(1:3, 5, 7:14)]
```

## Conclusions
Inspection of correlation and scatterplot matrices for the Boston data set 
revealed multiple linear associations between pairs of variables, as well as 
other noticeable trends; future analysis should carefully consider these 
correlations and perhaps eliminate them as to avoid multicollinearity.  

A particular trend was observed for plots of per capita crime rate (**crim**) 
against predictors **zn**, **indus**, **chas**, **rad**, **tax**, and 
**ptratio**, where the near entirety of the variation of **crim** was found at 
one specific predictor level.  Suburbs which shared this specific combination of 
predictor levels were found to comprise 93.1% of the suburbs in the top 25% with 
respect to per capita crime rate.  

Numerical summaries for suburbs having the minimum median household value 
(**medv** = 5) observed in the data suggested that lower values may be 
characterized via higher NO concentrations, older homes with relatively few 
rooms, a higher percentage of lower status population, and close proximities to 
Boston employment centers.  Summaries and t-tests for suburbs found to have
a high average number of rooms per dwelling (**rm** > 8) suggested that these
suburbs tend to score higher on measures for the predictor **black** and median
household value, and lower measures for per capita crime rate, degree of 
industrialization, property-tax rates, pupil-teacher ratios, and percentage of
population regarded as lower status.


## Appendix I: R Code

The following code does not generate any figures and supplements that which is 
found in the Plots and Tables section.

```{r Missing, results = "hide"}
which(is.na(df) == TRUE) # Locates missing values, if any; returns 0
```

```{r Chas, results = "hide"}
dim(df[df$chas == 1, ]) # [35 14] ==> 35 suburbs bound the Charles River
35/506 # 0.0692 = proportion of suburbs on Charles River
7/127 # 0.0551 = proportion of "high crime" (top 25%) suburbs on Charles River
```

```{r ChasTest}
# Conduct 2-sample proportion z-test at 99% confidence
off.river <- c(120, 471)
total <- c(127, 506)
prop.test(off.river, total, conf.level = 0.99)
```

```{r More8, results = "hide"}
dim(df[df$rm > 7, ]) # [64 14] ==> 64 suburbs average more than 7 rooms/dwelling
dim(df[df$rm > 8, ]) # [13 14] ==> 13 suburbs average more than 8 rooms/dwelling
```

